Understanding Factors Contributing to Application Approval in Urban Planning Decisions

### 2024-11-28 paper/paper.qmd
Upload Data part

library(tidyverse)
library(opendatatoronto)
library(modelsummary)

@@

Our data [@scommittee_application]....

The analyses presented in this paper were conducted using R programming language [@citeR]. The `tidyverse` packages [@citetidyverse], `dplyr` package [@citedplyr], and `arrow` package [@citearrow] were used in the process of data simulation, testing beforehand. After the original raw data was downloaded by using `tidyverse` package [@citetidyverse], and `dplyr` package [@citedplyr], data cleaning process was done by using `tidyverse` package [@citetidyverse], `dplyr` package [@citedplyr], and `arrow` package [@citearrow]. The code style is corrected by `here` package [@citehere], `lintr` package [@citelintr], and `styler` package [@citestyler]. Then, models were constructed using `tidyverse` package [@citetidyverse], and `rstanarm` [@citerstanarm] package. Graphs were made with `ggplot2` package [@citeggplot]. The model results are then presented by `modelsummary` [@citemodelsummary] package.

Following @tellingstories, we consider...

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

outcome variable: decision

@fig-decision shows the distribution of application decisions.

@@

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

Talk way more about it.

## Predictor variables

@@

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@fig-application-types presents the distribution of application types.

@@

@@

@fig-year presents the distribution of application years

@@

@@

@fig-district shows the distribution of planning district.

@@

@@

Here we briefly describe the Bayesian analysis model used to investigate, and include justification for model and the variables, as well as discuss underlying assumptions, potential limitations, software used to implement the model, and evidence of model validation and checking.

@@

### Model Implementation

@@

# Additional data details

>>>>

library(tidyverse)
library(opendatatoronto)
library(modelsummary)
library(knitr)

@@

Our data [@committee_application] is extracted from opendatatoronto [@opendatatoronto]. This dataset focuses on applications submitted to the City of Toronto's Committee of Adjustment, a governing body responsible for reviewing minor variance and consent applications related to land use and development. These applications are submitted by property owners and developers seeking permission to deviate from existing zoning regulations or to subdivide land. The Committee’s decisions—whether to approve or refuse an application—have significant implications for urban development, community planning, and city growth.

The Committee of Adjustment plays a key role in Toronto’s urban planning ecosystem. It ensures that deviations from zoning bylaws align with the city’s broader development objectives, maintaining a balance between flexibility for property owners and adherence to long-term planning goals. The decisions made by the committee are influenced by various factors, including the type of application, geographic considerations (e.g. planning districts), and the timing of submission. Understanding these decisions provides insight into urban planning dynamics and the regulatory challenges faced by a growing city like Toronto.

The analysis presented in this paper were conducted using R programming language [@citeR]. The `tidyverse` packages [@citetidyverse], `dplyr` package [@citedplyr], and `arrow` package [@citearrow] were used in the process of data simulation, testing beforehand. After the original raw data was downloaded by using `tidyverse` package [@citetidyverse], and `dplyr` package [@citedplyr], data cleaning process was done by using `tidyverse` package [@citetidyverse], `dplyr` package [@citedplyr], and `arrow` package [@citearrow]. The code style is corrected by `here` package [@citehere], `lintr` package [@citelintr], and `styler` package [@citestyler]. Then, models were constructed using `tidyverse` package [@citetidyverse], and `rstanarm` [@citerstanarm] package. Graphs were made with `ggplot2` package [@citeggplot]. Tables were constructed with `knitr` package [@citeknitr]. The model results are then presented by `modelsummary` [@citemodelsummary] package.

Following @tellingstories, this paper uses Bayesian Logistic Model to explore factors that affect the application results.

Although several similar datasets with overlapping themes were considered but ultimately not used due to limitations in scope, structure, or relevance. For instance, the Toronto Application Information Centre Dataset provides information on broader planning applications like rezoning requests and site plan approvals, but its complexity and inclusion of unrelated application types make it unsuitable for isolating patterns related to minor variances and consents. The Ontario Municipal Board (OMB) Decisions Dataset includes appeal outcomes, but its focus on contested cases introduces bias and omits the routine applications central to this study.

Consequently, focusing on the Committee of Adjustment Applications dataset [@committee_application] stands out for its granularity and relevance, enabling a focused analysis of approval trends and their influencing factors, which directly aligns with the study's goals.

@@

The dataset captures information on applications submitted to the City of Toronto’s Committee of Adjustment, providing a structured record of decisions made on zoning variances and consents. The dataset’s structure results from Toronto’s well-defined application process. When property owners or developers submit applications to the Committee of Adjustment, they provide detailed information about their project, including its location, purpose, and the specific variances or consents requested. These details are reviewed by committee staff and entered into a digital database, ensuring consistency and standardization.

Applications to the Committee of Adjustment are typically submitted by property owners, developers, or their authorized agents (e.g., architects or planners). Eligibility to submit an application requires ownership or legal rights to the property in question. For some application types, such as consents, additional prerequisites include demonstrating compliance with existing zoning bylaws or providing documentation showing that the proposed subdivision aligns with city planning objectives. Minor variance applications, on the other hand, require applicants to show that the requested adjustments are “minor” in nature, do not conflict with broader zoning intentions, and maintain compatibility with the surrounding area. As a result, individuals submitting these applications are often professionals with expertise in urban planning, real estate, or development, or property owners working in consultation with such professionals.

The data collection process begins with property owners or developers submitting applications through a centralized system. Each application includes essential details, which are systematically verified and recorded by administrative staff. Following public hearings, committee decisions (approval or refusal) are added to the database, completing the data entry process. The dataset is then maintained and periodically published by the City of Toronto, ensuring it remains accessible and accurate for public and analytical use.

The dataset thus provides a structured lens through which to examine urban planning decisions. By distilling real-world phenomena—such as zoning adjustments and land use approvals—into discrete variables, the dataset enables systematic analysis of trends and relationships within Toronto’s regulatory framework. This structured approach ensures that the dataset is both practical for analysis and representative of the nuanced processes underlying urban development.

## Data Cleaning

To ensure the dataset was suitable for analysis, several high-level cleaning steps were performed. First, unused decision outcomes were filtered out, retaining only applications that were explicitly approved or refused. The `decision` variable was then converted into a binary format, with `1` representing approvals and `0` representing refusals, simplifying its use in modeling.

A new variable, `year`, was constructed by extracting the year from the `in_date` field. This allowed for a clearer examination of temporal trends while maintaining the granularity of the original data. Applications from very early years with minimal counts were excluded to avoid bias and ensure a representative analysis of contemporary decision-making trends.

Additionally, key variables are selected out in cleaned dataset for convenience. More details will be explored in appendix.

## Features

The dataset provides details of applications submitted to the Committee of Adjustment, focusing on key variables:

- Decision: Indicates whether an application was approved or refused.
- Application Type: Categorizes the application as either a minor variance or a consent.
- Year: The year when the application was submitted.
- Planning District: The geographic area within Toronto where the application originated.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-preview
#| tbl-cap: "Preview of Key Variables"

# Show first six rows of the cleaned dataset
analysis_data |>
  head() |>
  kable(booktabs = TRUE, align = "c")
```

The first six rows of the cleaned dataset, displayed in @tbl-preview, offer a clear preview of these key variables.

These variables allow for an in-depth analysis of the factors influencing application decision results and provide a quantitative basis for evaluating trends in urban planning decisions. Specific and further discussion about these variables are included in @sec-data-outcome and @sec-data-predictor.

## Outcome variables {#sec-data-outcome}

The outcome variable, decision, captures whether an application submitted to the Committee of Adjustment was approved (represented as 1) or refused (represented as 0).

@@

@fig-decision illustrates the distribution of decision outcomes. Out of the total applications, 1,207 (66.5%) were approved, while 607 (33.5%) were refused. This demonstrates that the majority of applications submitted to the Committee of Adjustment are approved, suggesting that most requests align with zoning regulations or are deemed acceptable following the committee's review process.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-decision
#| tbl-cap: "Summary Statistics of Decisions"

#### Generate summary statistics for `decision` ####
decision_summary <- analysis_data %>%
  group_by(decision) %>%
  summarize(
    Percentage = paste0(round((n() / nrow(analysis_data)) * 100, 1), "%") # Add '%' symbol
  ) %>%
  mutate(Decision = ifelse(decision == 1, "Approved (1)", "Refused (0)")) %>%
  select(Decision, Percentage)

#### Create a kable table ####
decision_summary %>%
  kable(
    format = "html", 
    col.names = c("Decision", "Percentage")
  )
```

@tbl-decision provides a clear picture of the approval and refusal rates within the dataset, highlighting a relatively high likelihood of approval for submitted applications.

## Predictor variables {#sec-data-predictor}

@@

The application type variable categorizes each application submitted to the Committee of Adjustment into two types:

Consent (CO): Requests related to the subdivision of land into separate parcels, often for new developments or property sales.

Minor Variance (MV): Applications seeking small adjustments to zoning regulations, such as building setbacks, height limits, or lot coverage.

These categories represent distinct regulatory processes, with Consents generally requiring more extensive evaluation compared to Minor Variances due to their implications on land use.

@@

@fig-application-types presents the distribution of applications by type. Minor Variances account for the majority of applications which reach to 1371 (75.6%), reflecting the routine nature of these requests in urban development. In contrast, Consents are less frequent which only have 443 counts (24.4%), likely due to their more specific requirements and potential complexity.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-application-type
#| tbl-cap: "Summary Statistics of Application Types"

#### Summary Statistics for Application Type ####
application_summary <- analysis_data %>%
  group_by(application_type) %>%
  summarize(
    Percentage = paste0(round((n() / nrow(analysis_data)) * 100, 1), "%") # Add %
  ) %>%
  mutate(Application_Type = ifelse(application_type == "MV", "Minor Variance (MV)", "Consent (CO)")) %>%
  select(Application_Type, Percentage)

#### Create a kable table ####
application_summary %>%
  kable(
    format = "html", 
    col.names = c("Application Type", "Percentage")
  )
```

@tbl-application-type emphasize the dominance of Minor Variances in the dataset, providing context for their role in urban development.

The relationship between application_type and the outcome variable decision will be explored further to determine if approval rates differ between Minor Variances and Consents. Preliminary trends suggest that Minor Variances may have higher approval rates due to their less complex nature. Additional analyses may examine whether these patterns vary by planning district or year.

@@

The year variable represents the year of submission for each application in the dataset. It provides insight into temporal trends in applications submitted to the Committee of Adjustment. This variable is instrumental in identifying changes in the volume of applications over time, which may reflect policy shifts, urban development trends, or external factors like economic conditions.

@@

@fig-year clarifies the distribution of applications from 2016 to 2024, with notable fluctuations in annual counts. Fewer applications were submitted during 2016-2019, with counts declining steadily from 198 applications in 2016 to a low of 93 in 2019. This period may reflect a stabilization phase or lower activity in urban development processes. Afterwards, application counts increased significantly, peaking in 2023 with 378 applications, followed by a slight decline to 249 in 2024. The rise in 2023 may be due to heightened development activity or easing of restrictions from prior external events, such as the COVID-19 pandemic.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-year
#| tbl-cap: "Summary Statistics of Years"

#### Summary Statistics for Year ####
year_summary <- analysis_data %>%
  group_by(year) %>%
  summarize(
    Percentage = paste0(round((n() / nrow(analysis_data)) * 100, 1), "%")
  ) %>%
  arrange(year)

#### Create a kable table ####
year_summary %>%
  kable(
    format = "html", 
    col.names = c("Years", "Percentage")
  )
```

@tbl-year states how each year contributes to the total volume of applications.

The year variable is likely related to the decision outcome and application type. As approval rates may have varied over time due to changing city planning policies or committee priorities. Besides, temporal trends may reveal shifts in the proportion of Minor Variances (MV) versus Consents (CO), reflecting evolving development demands.

@@

The planning district variable identifies the geographic district within Toronto where an application was submitted. The city is divided into four planning districts, each with its own development characteristics and zoning regulations. These districts are crucial for understanding spatial trends in application submissions and outcomes, as local policies and priorities can influence decision-making processes.

@@

@fig-district displays the distribution of applications across four planning districts. We can observe that North York has the highest number of applications, accounting for a significant proportion of the dataset. This could reflect the area's high density of development projects or active property owners seeking zoning adjustments. Scarborough has the fewest applications, possibly due to fewer development activities or stricter zoning compliance. Etobicoke York and Toronto East York have moderate numbers of applications, reflecting their roles as active but less prominent districts compared to North York.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-district
#| tbl-cap: "Summary Statistics of Planning Districts"

#### Summary Statistics for Planning District ####
district_summary <- analysis_data %>%
  group_by(planning_district) %>%
  summarize(
    Percentage = paste0(round((n() / nrow(analysis_data)) * 100, 1), "%")
  )

#### Create a kable table ####
district_summary %>%
  kable(
    format = "html", 
    col.names = c("Planning Districts", "Percentage")
  )
```

@tbl-district provide a view of how applications are distributed spatially across the city.

The planning_district variable may also interacts with application type and decision. Approval rates may vary by district, reflecting differences in local policies or development dynamics. The proportion of Minor Variances (MV) versus Consents (CO) may differ across districts, indicating distinct development priorities or zoning challenges.

@@

Here we briefly describe the Bayesian analysis model used to investigate, and include justification for model and the variables, as well as discuss underlying assumptions, potential limitations, and evidence of model validation and checking.

@@

@@

# Additional data cleaning details

The dataset underwent a structured cleaning process to ensure its suitability for analysis, focusing on preparing the data while minimizing bias and enhancing interpretability. Below is a detailed description of the key cleaning steps:

1. **Filtering Unused Decision Outcomes**

The dataset originally included a range of decision outcomes beyond approvals and refusals, such as applications that were deferred or withdrawn. These were excluded to focus solely on cases where the Committee of Adjustment made a definitive decision. This exclusion is done by filtering decisions with "approve", "approval", and "refused" words in it while conditionally approved are all default as approval to enlarge the data size. This ensured a binary structure for the decision variable, aligning with the study’s objectives.

2. **Conversion of decision to Binary Format**

The decision variable was converted into a binary format for consistency and ease of analysis:

1: Representing approved applications.
0: Representing refused applications. This transformation provided a clear and standardized representation of application outcomes.

3. **Creation of the year Variable**

A new variable, year, was created by extracting the year component from the date field. This allowed for an analysis of temporal trends without losing the temporal resolution of the data. The original date variable remained intact for potential use in supplementary analyses.

4. **Filtering Applications from Earlier Years**
Applications before 2016 were removed from the dataset due to their low counts. Retaining these outliers could lead to biased results, particularly in trend analyses or modeling. By excluding these years, the dataset better represents the more recent and consistent patterns of decision-making.

5. **Removing Missing Values**
Records with missing values in the chosen variables (decision, application_type, year, and planning_district) were removed.
